\chapter{Literature Review}
\label{chap:literature}

This chapter surveys the five research traditions that the thesis
integrates: classical truss sizing, evolutionary and swarm
metaheuristics, neural surrogate modelling, reinforcement learning
for structural design, and large-language-model assisted engineering.
A final section summarises the Indian Standard~IS~800:2007 steel
design code and identifies the specific gap in the published
literature that this thesis addresses.

\section{Classical Truss Optimisation}
\label{sec:lit:classical}

Structural sizing optimisation has a continuous research
thread reaching back to the mid-1960s. The foundational
formulation of minimum-weight truss sizing under stress and
displacement constraints is due to Schmit and
Farshi~\cite{schmit1974}, who cast the problem as a constrained
non-linear programme and introduced the idea of
\emph{approximation concepts} --- replacing the exact implicit
dependence of constraints on design variables with explicit
first-order Taylor expansions at each design iterate. The
Schmit--Farshi 10-bar planar cantilever truss, introduced as an
illustrative example in that 1974 paper and refined in a 1976 NASA
contractor report~\cite{schmit1976}, is still the single most-cited
benchmark in the field and serves as the primary validation case in
Chapter~\ref{chap:results} of this thesis.

Venkayya~\cite{venkayya1971} gave an earlier, independent
optimality-criterion treatment, and his 25-bar spatial-transmission-
tower benchmark (two load cases, four symmetry groups) has become
the second canonical test case for sizing algorithms. Fleury and
Schmit~\cite{fleury1980} extended the approximation-concepts
framework to \emph{dual methods}, introducing the 72-bar
four-storey spatial tower as a more geometrically ambitious third
benchmark.

Trust-region variants of the Schmit--Fleury approach culminated in
Sunar and Belegundu's~\cite{sunar1991} exact second-order sensitivity
method, which reproduced the canonical 10-bar optimum of
\num{5060.85}~lb to four-decimal-place accuracy and set a standard
that every subsequent solver on this benchmark is compared against.
The methods of this era --- Schmit 1974 through Sunar 1991 --- are
all gradient-based: they assume differentiability of the constraint
functions and use either the dual-method optimality criterion or a
sequential-quadratic-programming (SQP) style inner solve. They are
exact, efficient on small problems, and remain the reference against
which later heuristic approaches are judged.

A shift toward population-based heuristic search is marked by
Erbatur et al.~\cite{erbatur2000}, whose genetic algorithm for
planar and spatial trusses produced near-optimal solutions on the
25-bar and 72-bar benchmarks without requiring gradient information.
The appeal of their approach is threefold: it handles discrete
design-variable sets (rolled-section catalogues) naturally; it is
insensitive to local minima; and it parallelises trivially across
population members. The price is that convergence is not exact,
and the weight of the best solution is typically within a few
percent of the gradient-based optimum rather than identical.

Camp and Bichon~\cite{camp2004} refined the 72-bar benchmark
formulation, explicitly listing symmetry groups, nodal
coordinates, two load cases, and a $\pm\num{0.25}~\mathrm{in}$
tip-deflection constraint; their paper is the canonical reference
for 72-bar encodings used by all subsequent ant-colony and
swarm-based papers. Bekdas\ et al.~\cite{bekdas2015} used the flower
pollination algorithm on the same four benchmarks and illustrate
the tendency of the field, post-\num{2000}, to proliferate
population-based heuristics of ever-more-exotic metaphor while
reporting improvements of a few tenths of a percent over the
earlier published values.

Two general observations frame this literature for the present
thesis. First, the \emph{canonical benchmarks} --- Schmit's 10-bar,
Venkayya's 25-bar, Fleury--Schmit's 72-bar, and Kaveh and
Talatahari's 200-bar~\cite{kaveh2010} --- have stabilised as the
de-facto validation set. Any new sizing algorithm is expected to
report on these before being taken seriously. Second, \emph{the
optimum weights are known to four to five significant figures} and
have been reproduced by multiple independent groups. This means a
new framework can be quantitatively validated: if a solver claims
to produce an optimum, it must land within a fraction of a percent
of the published number or be treated with suspicion. The
reproducibility standard set by the classical literature is what
makes Chapter~\ref{chap:results}'s \num{0.004}~\% match on 10-bar a
meaningful result rather than a rhetorical flourish.

\section{Evolutionary and Swarm Metaheuristics}
\label{sec:lit:metaheuristics}

Beyond the tailored genetic-algorithm formulations of
Erbatur~\cite{erbatur2000}, a general-purpose evolutionary-
computation literature has developed that this thesis draws on
directly. Kennedy and Eberhart~\cite{kennedy1995pso} introduced
particle swarm optimisation, motivated by an idealisation of
social-coordination dynamics in bird flocks. A PSO particle carries
a position, a velocity, a memory of its best personal visit, and
sees the swarm's best global visit; the velocity update is a
weighted pull toward both memories plus a stochastic perturbation.
PSO is attractive for truss sizing because it has few
hyperparameters (inertia, cognitive, social weights), converges
quickly on smooth landscapes, and requires only objective-function
evaluations --- it inherits the gradient-free advantage of GA
without GA's crossover-operator design choices. The PSO results in
Chapter~\ref{chap:results} use the default pymoo
hyperparameters~\cite{blank2020pymoo}, which are essentially the
Kennedy--Eberhart 1995 recommendations.

Deb~\cite{deb2001} wrote the standard textbook on multi-objective
optimisation using evolutionary algorithms. For problems with a
single objective (minimise weight subject to constraints) the
distinction is moot, but Chapter~\ref{chap:results} of this thesis
additionally reports weight-vs-maximum-displacement Pareto fronts
computed by the NSGA-II algorithm of Deb, Pratap, Agarwal, and
Meyarivan~\cite{deb2002nsga2}. NSGA-II introduced three
innovations that make it the workhorse of applied multi-objective
GA: fast non-dominated sorting in $O(MN^2)$ time (with $M$
objectives and $N$ population size), a crowding-distance metric
that preserves Pareto-front diversity without requiring
user-specified sharing parameters, and a rank-plus-crowding
selection operator that treats both as first-class. NSGA-II is the
default multi-objective solver in pymoo and is used unmodified in
this thesis.

Constraint handling in evolutionary algorithms is its own sub-
literature. Coello~\cite{coello2002} surveys the state of the art
as of the early 2000s and groups methods into penalty functions,
repair operators, feasibility-preserving decoders, and ranking-
based approaches. This thesis uses the static-penalty formulation
most common in truss-sizing literature: the fitness value is the
weight plus a large coefficient times the sum of squared
constraint violations. This is the default behaviour of pymoo's
constraint-handling interface and, on the benchmarks surveyed in
Chapter~\ref{chap:results}, consistently keeps the final feasible
population pressed against the active-constraint boundary.

The arrival of Blank and Deb's pymoo
library~\cite{blank2020pymoo} in 2020 is a convenience revolution
for applied optimisation. Pymoo exposes GA, PSO, NSGA-II,
NSGA-III, CMA-ES, and differential evolution under a uniform
\texttt{Problem--Algorithm--Termination} interface; switching
solver is a single-line change in the caller. The thesis's
\texttt{src/algorithms/} module is a thin Problem-subclass adapter
over pymoo's engine, and the default hyperparameter sets
(population \num{100}, \num{500} generations, simulated-binary
crossover with distribution index \num{15}, polynomial mutation,
tournament selection of size \num{2}) are used throughout.

\section{Neural Surrogate Models for Structural Analysis}
\label{sec:lit:surrogate}

A surrogate model is an inexpensive approximation to an expensive
simulator. In structural optimisation the expensive simulator is
the finite-element solve; the surrogate is some function --- a
polynomial response surface, a radial-basis network, a Kriging
Gaussian process, or a multi-layer perceptron --- trained on a
sample of FEM-evaluated design points and then used in place of
the FEM call inside the optimiser's inner loop.

The canonical survey of surrogate-based optimisation is Queipo et
al.~\cite{queipo2005}, who catalogue response-surface, Kriging,
radial-basis-function, and neural-network surrogates and lay out
the design-of-experiments sampling strategies (Latin hypercube,
orthogonal arrays, quasi-random sequences) that make surrogate
training data-efficient. Forrester, S\'obester, and
Keane~\cite{forrester2008} give a textbook treatment of Kriging
surrogates and adaptive infill criteria; their \emph{expected-
improvement} acquisition function is the foundation of modern
Bayesian optimisation. These two references establish the
methodological framework --- the sample, train, evaluate,
optionally adaptively refill loop --- that the thesis's
\texttt{src/ml/} module follows.

The recent trend is toward deep-learning surrogates trained on
Latin-hypercube-sampled datasets of FEM evaluations. Persia et
al.~\cite{persia2025} used MLP surrogates as a post-processing
accelerator in topology optimisation and reported two-order-of-
magnitude speedups with acceptable accuracy. Sun et
al.~\cite{sun2023} coupled a deep neural network surrogate with a
genetic algorithm for the optimisation of composite structures and
demonstrated that the surrogate--GA combination converged to
near-identical optima as the FEM--GA baseline while cutting wall-
clock time by roughly $50\times$.

The specific architecture used in Chapter~\ref{chap:methodology}'s
surrogate layer --- a three-hidden-layer MLP with widths
$[\num{256}, \num{128}, \num{64}]$, ReLU activation, dropout
\num{0.1}, Adam optimiser~\cite{kingma2015adam} at learning rate
\num{e-3}, and a multi-head output (weight, maximum-displacement,
maximum-member-stress) --- is a minimalist choice grounded in
Goodfellow, Bengio, and Courville's deep-learning
textbook~\cite{goodfellow2016dl}. The network has approximately
\num{50000} parameters and trains in under two minutes on a
\num{10000}-sample LHS dataset. Weight is learned almost exactly
($R^2 = \num{0.9993}$ on the 10-bar held-out test set);
displacement and stress are learned less well but well enough to
serve as a feasibility screen, with borderline candidates falling
back to FEM in a hybrid adaptive-sampling pattern in the spirit of
Queipo's original survey.

\section{Reinforcement Learning in Structural Design}
\label{sec:lit:rl}

Reinforcement learning, in contrast to supervised surrogate
learning, trains an agent to \emph{act} in a design environment:
the agent observes a partial design, takes an action that modifies
it, receives a reward, and iterates until a terminal design is
reached. Sutton and Barto~\cite{sutton2018rl} provide the
canonical textbook. For continuous action spaces the dominant
algorithm at the time of this writing is Proximal Policy
Optimization (PPO) by Schulman et al.~\cite{schulman2017ppo}, a
policy-gradient method with a clipped surrogate objective that
trades off the aggressive updates of trust-region policy
optimisation against implementation simplicity.

Applications to structural optimisation are recent but growing.
Hayashi and Ohsaki~\cite{hayashi2020rl} used a graph-neural-network
encoder and a deep-Q-network agent for binary-truss topology
optimisation under stress and displacement constraints, reporting
that the learned policy could solve previously-unseen instances
without re-training. Brown and Mueller~\cite{brown2022rl}
applied RL to continuum topology optimisation and emphasised the
inference-time speed advantage: a trained policy produces a design
in a single forward pass, making the method suitable for
interactive design exploration. Zhao et al.~\cite{zhao2021rl}
studied truss topology optimisation with discrete sizing
constraints and observed that on any single-instance benchmark a
tuned genetic algorithm typically outperforms a model-free RL
agent; RL's advantage, they argue, is generalisation across
instances rather than performance on a single instance. This
observation frames the interpretation of the PPO results in
Chapter~\ref{chap:results}: the agent's \num{16}~\% optimality gap
on the single 10-bar instance is not a failure of the method but
an instance of the Zhao trade-off, and the corresponding
advantage --- inference in \num{0.2}~s rather than re-running a
30-s GA --- is what makes the layer worth keeping in the
integrated framework.

Rochefort-Beaudoin and colleagues~\cite{rochefort2024sogym}
recently published \emph{SOgym}, a Gymnasium-compatible benchmark
environment for structural-optimisation RL. \emph{SOgym} is the
spiritual antecedent of the \texttt{TrussDesignEnv} defined in
Chapter~\ref{sec:method:rl}; its design conventions --- a dict-
valued observation space, a Box action space over normalised
design variables, reward shaping by a weighted sum of objective
improvement and constraint violation --- are followed closely by
this thesis.

\section{LLM-Assisted Engineering Design}
\label{sec:lit:llm}

The use of large language models as \emph{designers} --- asking a
model to propose engineering artefacts rather than describe them
--- is a research area less than three years old at the time of
this writing, but it is moving rapidly. Wei et
al.~\cite{wei2022cot} established that chain-of-thought prompting
--- asking the model to reason step by step before committing to
an answer --- improves performance on compositional-reasoning
tasks substantially. Chain-of-thought is now standard practice for
any LLM-mediated design task, including the truss-sizing prompt
used in Chapter~\ref{sec:method:llm}.

Two papers are direct precedents for the work in this thesis.
Geng et al.~\cite{geng2024} integrated a large language model with
the OpenSeesPy structural-analysis package, letting the LLM
generate complete Python scripts for linear and non-linear
structural problems; they found the LLM could produce working
code for canonical problems but required human supervision at the
parameter level. Makatura and colleagues~\cite{makatura2024}
surveyed the broader question of how large language models can
assist humans in design and manufacturing, and concluded that
the most productive deployment pattern is \emph{the LLM as
co-designer} --- not replacing the human engineer, and not
replacing the optimiser, but providing warm-start designs,
natural-language interfaces, and documentation support that shorten
the human-in-the-loop cycle.

The LLM layer in this thesis is a realisation of the Makatura
co-designer pattern specialised to the warm-start use case.
Anthropic's Claude API~\cite{anthropic-docs} is queried with a
structured prompt that describes the benchmark geometry, loading,
and design bounds; the model returns a JSON design vector that is
injected as one member of the initial GA population. The novelty
is not in the LLM itself --- Claude is used via the public API
without fine-tuning --- but in the demonstration that LLM-
generated designs are quantitatively useful warm-starts. On the
10-bar benchmark with \num{40} random seeds, LLM-seeded GA runs
converged to the literature optimum in \num{36.8}~\% fewer
generations than random-seeded runs, with paired-Wilcoxon
$p = \num{0.0046}$. This is the principal empirical contribution
of the thesis.

\section{Indian Standard IS~800:2007 Design Code}
\label{sec:lit:is800}

Every reported optimum in this thesis is checked against the
Bureau of Indian Standards code \emph{IS~800:2007 General
Construction in Steel --- Code of Practice}~\cite{is800-2007}.
IS~800:2007 replaced the long-standing working-stress edition of
\num{1984} with a limit-state-design framework aligned with
Eurocode and AISC practice. The clauses relevant to truss sizing
are:
\begin{itemize}
  \item \textbf{Clause~3.7 and 3.8} (effective length and
        slenderness ratios): maximum slenderness
        $\lambda_{\max}=\num{180}$ for compression members carrying
        loads resulting from dead and imposed loads, and
        $\lambda_{\max}=\num{400}$ for tension members;
  \item \textbf{Clause~5.6.1} (serviceability deflection limits):
        vertical deflection of trusses supporting brittle cladding
        limited to $L/\num{325}$, where $L$ is the span;
  \item \textbf{Clause~6.2} and \textbf{6.3} (design strength in
        tension due to yielding of the gross section and rupture
        of the critical section respectively), and the
        block-shear provisions of \textbf{Clause~6.4};
  \item \textbf{Clause~7.1} (compression-member design, including
        the column-curve buckling-reduction factor $\chi$ as a
        function of non-dimensional slenderness
        $\bar{\lambda}$).
\end{itemize}

Subramanian's textbook \emph{Design of Steel
Structures}~\cite{subramanian2008} and Duggal's \emph{Limit State
Design of Steel Structures}~\cite{duggal2014} are the two standard
Indian references for IS~800:2007 and were used for worked-example
cross-checks of the compliance module. The older handbook
SP~6(1):1964~\cite{sp6-1-1964}, though published under the
superseded working-stress regime, still provides the standard
rolled-section catalogue that fixes the lower bound of the design-
variable vectors; no modern replacement is in current use in
Indian practice.

The IS~800:2007 compliance layer in this thesis is implemented as
a pure-Python module (\texttt{src/constraints/is800\_checks.py})
that takes a design vector, the FEM-computed axial forces, and the
member lengths, and returns a vector of constraint values (one per
clause per member). It is unit-aware --- internally using SI
throughout --- and includes unit conversions for the imperial-unit
benchmarks in the literature. The module is exercised by a pytest
suite that verifies each clause against a worked example from
Subramanian or Duggal.

\section{Research Gap}
\label{sec:lit:gap}

The five research traditions reviewed above each have a mature
literature and each a clear application to truss sizing. The gap
the present thesis fills is \emph{integrative}: no published work
combines classical evolutionary optimisation (GA, PSO, NSGA-II), a
neural surrogate accelerator, a reinforcement-learning policy, an
LLM warm-start designer, and Indian-code (IS~800:2007) compliance
checking in a single open-source reproducible framework. The
closest prior work --- Geng et al.~\cite{geng2024}'s LLM +
OpenSeesPy integration and Sun et al.~\cite{sun2023}'s DNN + GA
coupling --- each integrate two of the five layers. The thesis's
contribution is to wire all five together, validate each layer
against a published benchmark, and demonstrate on four canonical
test cases that the integrated stack reproduces the classical
optima to fractions of a percent while offering a
$\num{50}$--$\num{150}\times$ inner-loop speedup and a
statistically significant warm-start benefit. The next chapter
describes the architecture of that stack in detail.
