\chapter{Methodology}
\label{chap:methodology}

This chapter describes the computational architecture of the
framework in the order a candidate design flows through it: the
finite-element foundation (\S\ref{sec:method:fem}); the benchmark
definitions used for validation (\S\ref{sec:method:benchmarks});
the classical evolutionary optimisers that drive the outer
population loop (\S\ref{sec:method:optimisers}); the neural
surrogate that replaces the expensive FEM call inside that loop
(\S\ref{sec:method:surrogate}); the reinforcement-learning agent
that learns a direct design-to-areas mapping
(\S\ref{sec:method:rl}); the large-language-model warm-start
designer (\S\ref{sec:method:llm}); the IS~800:2007 compliance
layer that every candidate must pass
(\S\ref{sec:method:is800}); and the thin Streamlit and FastAPI
layer that exposes the stack to a user
(\S\ref{sec:method:ui}). Each section states the layer's input--
output contract, references the corresponding source module, and
lists the validation gate against which the layer was tested in
Chapter~\ref{chap:results}.

\section{Framework Overview}
\label{sec:method:overview}

Figure~\ref{fig:arch} --- reproduced from \texttt{docs/thesis\_vision.md}
\S2 --- sketches the eight-layer stack. An incoming design vector
$\mathbf{x}\in\mathbb{R}^{n_\text{vars}}$ (member cross-sectional
areas, possibly after symmetry grouping) is passed to either (a)
the finite-element solver for an exact evaluation, (b) the neural
surrogate for a 50-to-150$\times$ faster approximate evaluation, or
(c) the LLM designer for a one-shot warm-start proposal that
bypasses the loop entirely. Every evaluated design is routed
through the IS~800:2007 compliance checker before being reported as
feasible. The eight layers communicate through three small data
classes defined in \texttt{src/algorithms/result\_schema.py}:
\texttt{OptimizationResult}, \texttt{BenchmarkEvaluation}, and
\texttt{ComplianceReport}.

Three invariants hold across the stack. First, the design-variable
bounds are set once by the benchmark module
(\texttt{src/benchmarks/truss\_10bar.py} and sibling files) and
every downstream layer reads them from that single source.
Second, units are SI internally; the benchmark definitions carry
an explicit unit tag and are converted to SI on load, so the
IS~800:2007 module never sees imperial inputs. Third, all randomness
is seeded through a single \texttt{seed} argument propagated
from the top-level CLI scripts (\texttt{scripts/run\_single.py},
\texttt{scripts/run\_batch.py}) into numpy, pymoo, and PyTorch,
making every result in Chapter~\ref{chap:results} exactly
reproducible.

\section{Finite-Element Foundation}
\label{sec:method:fem}

The finite-element engine is a textbook pin-jointed truss solver
written from scratch in \texttt{src/fem/}. It is deliberately
minimal --- two-node linear bar elements, direct-stiffness assembly,
and a partition-method boundary-condition solve --- because the
thesis's thesis is about the \emph{optimiser} layer, not a new FEM
formulation. The implementation follows
Logan~\cite{logan2017fem} and Cook et al.~\cite{cook2002fem}
chapter by chapter.

\paragraph{Element stiffness.} For a bar element with nodes $i$ and
$j$, length $L$, cross-section $A$, Young's modulus $E$, and
direction cosines $c=\cos\theta$, $s=\sin\theta$ (extended to
three dimensions for spatial trusses), the $4\times 4$ (2D) or
$6\times 6$ (3D) element stiffness in global coordinates is
\begin{equation}
    \mathbf{k}_e = \frac{AE}{L}\,\mathbf{T}^{\!\top}
    \begin{bmatrix} 1 & -1 \\ -1 & 1 \end{bmatrix} \mathbf{T},
    \label{eq:elem_stiff}
\end{equation}
where $\mathbf{T}$ is the $2\times 4$ (or $2\times 6$) transformation
that projects global nodal displacements onto the bar's axial
direction. Equation~\eqref{eq:elem_stiff} is implemented once in
\texttt{src/fem/truss\_element.py} and handles both the 2D
(10-bar, 200-bar) and 3D (25-bar, 72-bar) benchmarks from a
single code path.

\paragraph{Global assembly.} Element stiffnesses are summed into a
global matrix $\mathbf{K}$ of size
$(n_\text{DOF}\times n_\text{DOF})$ by adding each
$\mathbf{k}_e$ into the rows and columns picked out by its node
list; numpy's \texttt{ix\_} indexing does the scatter in one
line (\texttt{src/fem/assembly.py}). Before any boundary
conditions are applied $\mathbf{K}$ is symmetric but singular, as
the structure is still free to translate and rotate as a rigid body.

\paragraph{Boundary conditions and solve.} The partition method
splits the global DOF set into \emph{free} and \emph{prescribed}
subsets, solving
\begin{equation}
    \mathbf{K}_{ff}\,\mathbf{u}_{f} =
    \mathbf{F}_{f} - \mathbf{K}_{fp}\,\mathbf{u}_{p}
    \label{eq:partitioned_solve}
\end{equation}
for the unknown free-DOF displacements $\mathbf{u}_f$, then
back-substitutes to recover the reactions.
Equation~\eqref{eq:partitioned_solve} is solved with
\texttt{scipy.linalg.solve} on the dense partitioned matrix;
the benchmark systems are at most \num{200} DOFs so sparse solvers
are unnecessary. Post-processing (axial forces, member stresses,
nodal displacement magnitudes) is performed in
\texttt{src/fem/post\_process.py}.

\paragraph{Validation gate.} The canonical three-bar truss of
Logan (Example~3.5) has a known analytical solution.
\texttt{tests/test\_fem.py} runs the three-bar case and asserts
relative error below $\num{e-12}$ on both nodal displacement and
axial force; the observed error is approximately $\num{e-21}$, i.e.\
at the floating-point epsilon of double-precision arithmetic. This
test must pass before any downstream benchmark is evaluated.

\section{Benchmark Truss Problems}
\label{sec:method:benchmarks}

Four canonical benchmarks from the literature serve as validation
fixtures. Each is encoded as a subclass of
\texttt{BenchmarkProblem} in \texttt{src/benchmarks/}, which
bundles node coordinates, element connectivity, material
properties, load cases, support conditions, design-variable
bounds, symmetry groups, and the published optimum area vector
and weight. Table~\ref{tab:benchmarks} summarises the four.

\begin{table}[htbp]
  \centering
  \caption{Canonical benchmark trusses.}
  \label{tab:benchmarks}
  \begin{tabular}{lcccc}
    \toprule
    Benchmark & Dim & Bars & Groups & Lit.\ optimum \\
    \midrule
    10-bar cantilever \cite{sunar1991}   & 2D & 10  & 10 (none) & \num{5060.85} lb \\
    25-bar tower \cite{venkayya1971}     & 3D & 25  & 8         & \num{545.22} lb  \\
    72-bar tower \cite{fleury1980}       & 3D & 72  & 16        & \num{379.62} lb  \\
    200-bar planar \cite{kaveh2010}      & 2D & 200 & 29        & \num{25445} lb   \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{10-bar cantilever (Sunar 1991).} Two-dimensional, six
nodes, ten bars, two vertical point loads of $-\num{100}$~kips at
the two lower free nodes, steel ($E = \num{e7}$ psi,
$\rho = \num{0.1}$ lb/in\textsuperscript{3}). Design variables are
the ten cross-section areas with bounds $[\num{0.1},
\num{35}]$~in\textsuperscript{2}. No symmetry grouping. The 10-bar
is the thesis's primary validation case because it is the
most-cited benchmark in the field and because its optimum is known
to four decimal places. Chapter~\ref{chap:results} reproduces the
Sunar 1991 optimum to $\num{0.004}$~\% (PSO) and $\num{0.04}$~\%
(GA).

\paragraph{25-bar spatial tower (Venkayya 1971).} Three-dimensional,
ten nodes, twenty-five bars, two load cases, aluminium-alloy
properties ($E = \num{e7}$ psi, $\rho = \num{0.1}$
lb/in\textsuperscript{3}). Design variables reduce from twenty-five
to eight by the symmetry grouping listed in
Venkayya's paper. Our GA and PSO reach \num{545.30} and \num{545.26}
lb respectively against the \num{545.22} lb literature value.

\paragraph{72-bar spatial tower (Fleury \& Schmit 1980).}
Three-dimensional, twenty nodes, seventy-two bars, two load cases,
sixteen symmetry groups per Camp and Bichon~\cite{camp2004}.
Chapter~\ref{chap:results} documents a residual optimality gap on
this benchmark; the investigation in \S\ref{sec:results:72bar}
concludes that the published \num{379.62} lb figure is infeasible
under strict FEM+IS~800 re-analysis at approximately
$\num{25}$~\% constraint violation, consistent with the long-
standing observation in the heuristic-optimisation literature that
many 72-bar ``optima'' rely on softened constraint tolerances.

\paragraph{200-bar planar (Kaveh \& Talatahari 2010).} Two-
dimensional, seventy-seven nodes, two hundred bars, twenty-nine
symmetry groups, multi-load case. Encoded as a stubbed benchmark
in \texttt{src/benchmarks/truss\_200bar.py}; full validation is
flagged as future work in Chapter~\ref{chap:conclusion}.

Each benchmark exposes a uniform interface: \texttt{evaluate(x)}
returns a \texttt{BenchmarkEvaluation} dataclass containing total
weight, maximum member stress, maximum nodal displacement, and
per-member feasibility flags. Optimisers, surrogates, and the
Streamlit UI all consume this single interface.

\section{Classical Optimisation Wrappers}
\label{sec:method:optimisers}

The classical optimiser layer wraps pymoo~\cite{blank2020pymoo}
behind a uniform \texttt{TrussProblem} adapter
(\texttt{src/algorithms/problem.py}). The adapter exposes two
methods: \texttt{\_evaluate(x, out)} which computes the objective
(total weight) and constraint-violation vector, and
\texttt{\_get\_var\_bounds()} which returns the lower and upper
design-variable bounds from the benchmark definition. The
constraint-violation vector collects IS~800:2007 stress, deflection,
and slenderness residuals; feasibility is defined as every element
of that vector being $\le 0$.

\paragraph{Genetic algorithm.} \texttt{src/algorithms/ga.py} calls
pymoo's GA with default operators: simulated-binary crossover
(SBX) with distribution index $\eta_c=\num{15}$ and probability
$p_c=\num{0.9}$, polynomial mutation with $\eta_m=\num{20}$ and
$p_m = 1/n_\text{vars}$, and tournament selection of
size \num{2}. Population size is \num{100} and generation budget is
\num{500}.

\paragraph{Particle swarm optimisation.}
\texttt{src/algorithms/pso.py} uses pymoo's canonical PSO with
inertia weight $w = \num{0.9}$ (linearly decreased to
$\num{0.4}$ over the run), cognitive weight $c_1 = \num{2.0}$, and
social weight $c_2 = \num{2.0}$, following
Kennedy~\cite{kennedy1995pso}. Swarm size and generation budget
match the GA settings.

\paragraph{NSGA-II.} \texttt{src/algorithms/nsga2.py} adapts the
same \texttt{TrussProblem} for the bi-objective case where the
second objective is maximum nodal displacement. NSGA-II's fast
non-dominated sort and crowding-distance selection run with the
same SBX/polynomial operators as the GA. The chapter-4 results
report a weight-vs-displacement Pareto front with at least twenty
non-dominated points on each benchmark.

\paragraph{Constraint handling.} All three solvers use the same
static-penalty formulation: the objective presented to pymoo is
$W(\mathbf{x})$ directly, and the constraint-violation vector
$g(\mathbf{x})$ is handled by pymoo's built-in feasibility-first
selection~\cite{coello2002}. Feasible individuals dominate
infeasible ones regardless of weight; among infeasibles the one
with smaller total violation dominates.

\paragraph{Runner and result schema.}
\texttt{src/algorithms/runner.py} is the single entry point that
takes a benchmark name, algorithm name, seed, and optional warm-
start vector $\mathbf{x}_0$, runs pymoo, and writes a populated
\texttt{OptimizationResult} to disk. The result schema
(\texttt{src/algorithms/result\_schema.py}) is a frozen
\texttt{dataclass} carrying the best area vector, best weight,
full population history, wall-clock time, and a compliance report
for the reported optimum.

\paragraph{Validation gate.} Every reported best design must
satisfy the IS~800:2007 compliance checker
(\S\ref{sec:method:is800}) under strict tolerances; every
benchmark's \num{10}-seed mean best-weight must land within
$\num{2}~\%$ of the literature optimum. Both gates pass on 10-bar
and 25-bar; 72-bar is discussed separately in
Chapter~\ref{chap:results}.

\section{Neural Surrogate Model}
\label{sec:method:surrogate}

The surrogate layer (\texttt{src/ml/}) replaces the FEM call with
a multi-layer perceptron. The motivation, as discussed in
\S\ref{sec:lit:surrogate}, is that a single 500-generation GA run
evaluates \num{50000} designs; at \num{50}~ms per FEM solve that is
forty minutes of wall clock, and the surrogate brings it below one
minute.

\paragraph{Architecture.} \texttt{src/ml/model.py} defines
\texttt{MLP}, a three-hidden-layer network with widths
$[\num{256}, \num{128}, \num{64}]$, ReLU activation,
dropout~\cite{goodfellow2016dl} at $p=\num{0.1}$, and a linear
output head of dimension three: $\log(1+W)$,
$\log(1+\delta_\text{max})$, and $\sigma_\text{max}$.
$\log(1+\cdot)$ transforms stabilise training on the long-tailed
weight and displacement distributions.

\paragraph{Dataset.} \texttt{src/ml/dataset.py} implements Latin-
hypercube sampling (LHS) via \texttt{scipy.stats.qmc} to draw
\num{10000} design vectors uniformly within the benchmark's
design-variable bounds. Each sample is evaluated by the FEM
engine and the triple $(W, \delta_\text{max}, \sigma_\text{max})$
stored alongside the input vector. LHS is preferred over
simple-random sampling because it guarantees one-dimensional
marginals are stratified, which in practice gives better coverage
of the extreme corners of the design space where the thin-
member optima tend to lie.

\paragraph{Training.} \texttt{src/ml/train.py} splits the LHS
dataset $\num{80}/\num{10}/\num{10}$ for train/validation/test.
Inputs and outputs are standardised to zero mean and unit
variance on the training split. The network is trained with the
Adam optimiser~\cite{kingma2015adam} at initial learning rate
$\num{e-3}$ under a cosine-annealing schedule, mini-batch size
\num{128}, and a \num{200}-epoch budget with early stopping on
validation mean-squared error.

\paragraph{Wrapped evaluator.} \texttt{src/ml/surrogate.py}
exposes a \texttt{SurrogateEvaluator} class with the same
\texttt{evaluate(x)} signature as \texttt{BenchmarkProblem},
making it a drop-in replacement inside the pymoo inner loop.
Inference is batched when pymoo passes a population.

\paragraph{Validation gate.} \texttt{src/ml/evaluate.py} runs the
trained model on the held-out test set, reports $R^2$ per output
head, and benchmarks inference throughput against FEM. On 10-bar,
$R^2_\text{weight} = \num{0.9993}$, $R^2_{\delta} = \num{0.89}$,
$R^2_{\sigma} = \num{0.82}$; FEM averages \num{41}~ms per design
against a surrogate \num{0.31}~ms per design (batched), a
$\num{132}\times$ speedup. The weight-$R^2$ figure is the target
gate from \S\ref{sec:intro:objectives}, reached with margin; the
lower displacement and stress $R^2$ values are interpreted as
feasibility screens with borderline candidates falling back to FEM
in the hybrid pattern of Queipo et al.~\cite{queipo2005}.

\section{Reinforcement-Learning Agent}
\label{sec:method:rl}

The reinforcement-learning layer (\texttt{src/rl/}) trains a
Proximal-Policy-Optimisation~\cite{schulman2017ppo} agent to
\emph{propose} designs directly, skipping the evolutionary outer
loop.

\paragraph{Environment.} \texttt{src/rl/environment.py} defines
\texttt{TrussDesignEnv}, a Gymnasium environment with
observation space equal to the normalised benchmark description
(bounds, load vector, symmetry groups packed into a fixed-length
tensor) and a Box action space of dimension $n_\text{vars}$ in
$[-1, 1]$, mapped to the design-variable bounds. The environment
is a single-step \emph{bandit} formulation: the agent produces a
design in one shot, the surrogate evaluates it, and the reward is
\begin{equation}
    r(\mathbf{x}) = -W(\mathbf{x}) - \lambda
    \sum_i \max(g_i(\mathbf{x}), 0)^2,
    \label{eq:rl_reward}
\end{equation}
with $\lambda=\num{e4}$. The quadratic penalty, rather than a
linear one, pushes the agent's proposals back \emph{inside} the
feasible region rather than sitting on the constraint boundary
where the surrogate is least accurate.

\paragraph{Training.} \texttt{src/rl/train\_ppo.py} uses Stable
Baselines3~\cite{schulman2017ppo}'s \texttt{PPO} class with
default hyperparameters (GAE $\lambda = \num{0.95}$, clip
$\epsilon = \num{0.2}$, learning rate $\num{3e-4}$) and
a rollout of \num{200000} environment steps. Because the
environment is single-step, each PPO update sees a fresh batch of
independent design proposals, effectively making this a contextual
bandit rather than a true sequential decision process. Training
takes approximately twenty minutes on a laptop CPU.

\paragraph{Evaluation.} \texttt{src/rl/evaluate.py} rolls the
trained policy for one deterministic rollout plus
$n_\text{rollouts}-1$ stochastic rollouts, re-evaluates each
proposed design with the exact FEM engine, and reports the best
feasible weight. The stochastic rollouts explore the boundary
layer around the policy mean, which empirically beats greedy
deterministic rollouts because the surrogate's approximation
error is largest exactly where the optimum sits.

\paragraph{Validation gate.} On 10-bar the PPO agent converges to
\num{5876.9}~lb, a \num{16.1}~\% gap against the
\num{5060.85}~lb literature value. This is consistent with
Zhao et al.~\cite{zhao2021rl}'s observation that model-free RL on a
single-instance structural problem does not generally match a
tuned GA on that same instance; the contribution of the RL layer
in this thesis is inference speed (approximately \num{0.2}~s per
design versus \num{30}~s for a fresh GA run), not single-
instance optimality.

\section{LLM Warm-Start Designer}
\label{sec:method:llm}

The LLM layer (\texttt{src/llm/}) queries Anthropic's Claude Opus
API~\cite{anthropic-docs} for a one-shot warm-start design that is
injected into the GA's initial population.

\paragraph{Prompt template.} \texttt{src/llm/prompts.py} builds a
system prompt (role description: ``you are a structural engineer
sizing an optimum-weight truss'') and a user prompt containing:
the benchmark name, the node coordinates in a compact tabular
form, the element list and loads, the material properties, the
IS~800:2007 clause citations to respect, and a request for a JSON
object of the form \texttt{\{"areas": [...]\}} with a chain-of-
thought field containing the model's working~\cite{wei2022cot}.

\paragraph{Client.} \texttt{src/llm/client.py} wraps the
\texttt{anthropic} Python SDK with two additions: an on-disk JSON
cache keyed by a content hash of the prompt (so re-running an
experiment with the same prompt does not re-bill) and a heuristic
fallback that returns a fully-stressed-design vector
$A_i = |F_i| / (\num{0.75}\,\sigma_\text{allow})$ when the API is
unavailable or the response fails to parse. The heuristic lets
the offline CI run and gives the pipeline graceful degradation.

\paragraph{Designer entry point.} \texttt{src/llm/designer.py}
exposes \texttt{suggest\_initial\_design(benchmark, seed)},
which builds the prompt, calls the cached client, parses the JSON,
clips to the design bounds, and returns the vector ready for pymoo.

\paragraph{Evaluation protocol.} \texttt{src/llm/evaluate.py}'s
\texttt{compare\_warmstart\_vs\_random(n\_seeds)} runs two
arms --- GA with LLM warm-start and GA with random initialisation
--- across $n_\text{seeds}$ seeds. For each run it records the
number of generations until the best feasible weight comes within
$\num{1.0}~\%$ of the literature optimum. A paired Wilcoxon
signed-rank test on the matched seed pairs reports the reduction
and its significance.

\paragraph{Validation gate.} On 10-bar with \num{40} seeds, LLM-
warm-started GA converges in a mean of \num{128.3} generations
against \num{203.1} for random initialisation, a reduction of
\num{36.8}~\%, with paired-Wilcoxon $p = \num{0.0046}$. This
clears the $\num{20}$--$\num{30}~\%$ gate in
\S\ref{sec:intro:objectives}.

\section{IS~800:2007 Compliance Layer}
\label{sec:method:is800}

Every reported optimum is checked against IS~800:2007 by the
compliance module (\texttt{src/constraints/}). Two files cooperate:
\texttt{is800\_checks.py} implements the individual clauses as
pure functions; \texttt{compliance.py} orchestrates them into a
\texttt{ComplianceReport} for a given design.

\paragraph{Clause 3.8 slenderness.} For each member, the
slenderness $\lambda = L_\text{eff}/r$ (with $r = \sqrt{I/A}$ the
radius of gyration, approximated for generic rod sections by
$r \approx \sqrt{A/\pi}$) must satisfy
$\lambda \le \num{180}$ in compression, $\lambda \le \num{400}$ in
tension. The sign of the axial force distinguishes the two cases;
the compliance module re-runs the FEM to extract signed forces
rather than the absolute-valued ones returned by the benchmark
\texttt{evaluate}.

\paragraph{Clause 6.2/6.3 tension.} Design strength in tension is
the minimum of yielding of the gross section
$T_{dg} = A_g f_y / \gamma_{m0}$ (Clause 6.2) and rupture of the
critical section $T_{dn} = \num{0.9} A_n f_u / \gamma_{m1}$
(Clause 6.3), with $\gamma_{m0}=\num{1.1}$ and
$\gamma_{m1}=\num{1.25}$ per the partial-safety-factor table. For
the benchmarks considered --- all pin-jointed rod members with no
hole deductions --- $A_n=A_g$, so the governing case is
typically Clause 6.3.

\paragraph{Clause 7.1 compression.} The design compressive stress
$f_{cd} = \chi\,f_y / \gamma_{m0}$ uses the column-curve reduction
factor $\chi$ as a function of non-dimensional slenderness
$\bar{\lambda} = \sqrt{f_y / f_{cc}}$ with
$f_{cc} = \pi^2 E / \lambda^2$ the Euler buckling stress. The
module implements the IS~800 column curve $a$ (the recommended
curve for rolled I-sections and hollow sections) using the
imperfection factor $\alpha = \num{0.21}$.

\paragraph{Clause 5.6.1 deflection.} Maximum nodal displacement is
limited to $L/\num{325}$ for trusses supporting brittle cladding.
For the cantilever and tower benchmarks $L$ is taken as the span
(10-bar) or the tower height (25-bar, 72-bar).

\paragraph{Unit conversions.} The benchmark optima are reported in
imperial units in the original papers; the IS~800 module operates
in SI. Conversions happen at the benchmark boundary
(\texttt{BenchmarkProblem.as\_si()}), not inside the compliance
checks, so the clauses above are written once and read SI inputs
always.

\paragraph{Validation gate.} \texttt{tests/test\_compliance.py}
exercises each clause against a worked example from
Subramanian~\cite{subramanian2008} or
Duggal~\cite{duggal2014}; all cross-checks agree to three
significant figures, the quoted precision of the textbook
solutions.

\section{User Interface and API}
\label{sec:method:ui}

The framework is exposed through two complementary entry points
living in \texttt{src/app/}.

\paragraph{FastAPI backend.} \texttt{src/app/api.py} defines a
REST API with endpoints \texttt{/health},
\texttt{/benchmarks}, \texttt{/benchmarks/\{name\}},
\texttt{/optimize}, and \texttt{/llm/suggest}. The
\texttt{/optimize} endpoint accepts a benchmark name, algorithm
(\texttt{ga}, \texttt{pso}, \texttt{nsga2}), generations, seed, and
optional LLM warm-start flag; it returns the best design, best
weight, convergence history, and compliance report in JSON. The
LLM warm-start is injected only for single-objective solvers
(GA, PSO) because pymoo's NSGA-II does not accept an initial-
population argument in the same form.

\paragraph{Streamlit UI.} \texttt{src/app/ui.py} is a thin client
over the FastAPI backend. It lets the user pick a benchmark and
algorithm, toggle LLM warm-start, set seed and generation count,
and visualise the result: a convergence curve, a Pareto front
(for NSGA-II), and a matplotlib rendering of the truss geometry
with member cross-sections drawn at scale.

\paragraph{Demo latency gate.} The end-to-end path
(UI $\to$ API $\to$ GA run $\to$ compliance report $\to$ JSON
back $\to$ UI render) completes in under ten seconds on 10-bar on
a laptop CPU, meeting the interactivity target in
\S\ref{sec:intro:objectives}. This is enabled by the surrogate
evaluator being wired in as an optional \texttt{use\_surrogate=true}
query parameter; without it the FEM inner loop caps throughput at
roughly forty seconds per run.

\paragraph{Module summary.} Table~\ref{tab:modules} lists the
source modules against the research objectives they implement.

\begin{table}[htbp]
  \centering
  \caption{Source modules against research objectives.}
  \label{tab:modules}
  \begin{tabular}{ll}
    \toprule
    Module & Implements objective \\
    \midrule
    \texttt{src/benchmarks/}   & Benchmark reproducibility (\S\ref{sec:intro:objectives}, obj.~1) \\
    \texttt{src/algorithms/}   & Classical optimiser accuracy (obj.~2) \\
    \texttt{src/ml/}           & Surrogate speedup + fidelity (obj.~3) \\
    \texttt{src/rl/}           & Reinforcement-learning feasibility (obj.~4) \\
    \texttt{src/llm/}          & LLM warm-start benefit (obj.~5) \\
    \texttt{src/constraints/}  & IS~800:2007 compliance (obj.~6) \\
    \texttt{src/app/}          & Streamlit demo \mbox{$<\num{10}$~s}  (obj.~6) \\
    \bottomrule
  \end{tabular}
\end{table}

The validation gates summarised at the end of each section above
are exercised in \texttt{tests/}, where a pytest suite of \num{76}
tests covers unit, integration, and gate-level assertions. The
results of running each gate against its benchmark are presented
in Chapter~\ref{chap:results}.
