\chapter{Conclusions and Future Work}
\label{chap:conclusion}

This final chapter collects the thesis's contributions, matches
them back against the research objectives of
\S\ref{sec:intro:objectives}, states the limitations honestly,
and maps out a research programme for the successor M.Tech or
Ph.D.\ cohort who picks up the codebase.

\section{Summary of Contributions}
\label{sec:concl:summary}

The thesis delivers five concrete, numerically validated
contributions.

\paragraph{C1. A validated integrated framework.} An end-to-end
Python framework (\texttt{src/} $+$ \texttt{scripts/} $+$
\texttt{thesis\_writeup/}) that composes seven previously-separate
research threads --- classical evolutionary optimisation
(GA/PSO/NSGA-II), finite-element analysis, neural surrogates,
reinforcement learning, LLM warm-start, IS~800:2007 compliance
checking, and a Streamlit+FastAPI user interface --- into one
reproducible system. The PSO-best on the 10-bar benchmark
matches Sunar \& Belegundu's 1991 published optimum
(\num{5060.85}~lb) to \num{5061.05}~lb, a $\num{0.004}\,\%$
error that acts as the end-to-end validation signature for the
entire stack
(\cref{sec:results:10bar}).

\paragraph{C2. A 100$\times$ neural surrogate.} A three-hidden-
layer MLP surrogate (widths $[\num{256},\num{128},\num{64}]$,
trained on \num{10000} Latin-hypercube samples)
achieves $R^2=\num{0.9993}$ on the weight head of the 10-bar
benchmark and delivers a $\num{132}\times$ wall-clock speedup
when used as a drop-in FEM replacement in the pymoo inner loop,
with surrogate-in-loop optima matching FEM-in-loop optima to
$\num{0.09}\,\%$
(\cref{sec:results:surrogate},
\cref{sec:results:ablation:surrogate}).

\paragraph{C3. A novel 72-bar infeasibility finding.} A
rigorous hard-constraint ($g \le 0$) re-run of the 72-bar
spatial tower exposes that the widely-cited literature optima
--- Camp \&~Bichon 2004 at \num{380.24}~lb, Bekdas et al.~2015
at \num{379.62}~lb --- are, under tight constraint enforcement,
infeasible by \num{22}--\num{62}\,\% on displacement and
\num{22}\,\% on stress. The true rigorously-feasible optimum
sits near $\num{549}$~lb. This is a methodological correction
to the published literature, not an optimiser shortfall; the
ablation of
\cref{sec:results:ablation:constraints} makes the point in
both figures and numbers
(\cref{sec:results:72bar}).

\paragraph{C4. An LLM warm-start efficacy map.} On the smallest
benchmark (10-bar, \num{40} seeds), an LLM warm-start reduces
generations-to-convergence by $\num{36.8}~\%$ against random
initialisation at paired-Wilcoxon $p=\num{0.0046}$ --- a strong
positive result driven by the LLM's ability to identify the
redundant-member set $\{2, 5, 6, 10\}$. On the two larger 3D
benchmarks the effect weakens: $\num{76.6}\,\%$ generation
reduction on 25-bar ($p=\num{0.25}$ with only three seeds) and
a null result on 72-bar ($-\num{4.8}~\%$ at $p=\num{0.81}$,
five seeds). The result maps out where LLM assistance helps in
the structural-optimisation pipeline: it helps most where
the LLM can surface an architectural insight --- redundant
members, symmetry groups --- that the GA would otherwise
rediscover through trial and error
(\cref{sec:results:llm}).

\paragraph{C5. An IS~800:2007-compliant, reproducible design
aid.} The Streamlit UI of \cref{sec:method:ui} exposes the full
stack --- benchmark selection, algorithm choice, surrogate
toggle, LLM warm-start toggle, IS~800 compliance report --- as
a single-command web app that a civil-engineering practitioner
can drive without writing Python. The complete pipeline
including IS~800 checking and a confidence-ranked
LLM-suggested cross-check finishes in under ten seconds on
10-bar on a laptop CPU. The repository is pinned to
\texttt{tag: phase-10-complete}; every reported number is
reproducible against that tag without an Anthropic API key
because all LLM responses are cached under
\texttt{results/llm\_cache/}.

\section{Answers to the Research Objectives}
\label{sec:concl:objectives}

Each of the five research objectives stated in
\S\ref{sec:intro:objectives} is answered below, with the
relevant result section cited in parentheses.

\paragraph{Objective 1 --- Reproduce classical benchmark
optima within \num{2}~\% of published values.} Achieved on
10-bar ($\num{0.004}\,\%$ error against Sunar~\cite{sunar1991})
and 25-bar ($\num{0.008}\,\%$ error against
Schmit~\cite{schmit1976}). On 72-bar the $\num{2}\,\%$ target
is unachievable against the published \num{379.62}~lb because
that optimum is infeasible under rigorous constraint
enforcement; the rigorous-feasibility optimum of $\num{549}$~lb
is itself reached to within $\num{0.13}~\%$ cross-seed
(\cref{sec:results:10bar,sec:results:25bar,sec:results:72bar}).

\paragraph{Objective 2 --- Generate Pareto fronts with at
least 20 non-dominated points.} Achieved: NSGA-II on 10-bar
returns \num{24}--\num{32} non-dominated points per seed
(\cref{sec:results:pareto}).

\paragraph{Objective 3 --- Achieve at least \num{50}$\times$
speedup via surrogate with $R^2 > \num{0.98}$.} Achieved
on the weight head ($R^2=\num{0.9993}$, $\num{132}\times$
speedup); partially achieved on the stress and displacement
heads ($R^2 = \num{0.81}/\num{0.87}$, used as feasibility
screens with FEM fallback, after
Queipo~\cite{queipo2005})
(\cref{sec:results:surrogate,sec:results:cost}).

\paragraph{Objective 4 --- PPO agent within \num{5}\,\% of the
best classical solver.} Not achieved in terms of single-
instance optimality ($+\num{16.12}\,\%$ gap on 10-bar); the
value of the layer is instead its $\num{0.2}$-s inference
latency, consistent with Zhao~\cite{zhao2021rl}'s
transferability framing. This is reported as a mixed result
rather than a claimed success
(\cref{sec:results:rl,sec:results:threats}).

\paragraph{Objective 5 --- LLM warm-start reduces generations-
to-convergence by at least \num{20}\,\%.} Achieved on 10-bar
($\num{36.8}\,\%$ reduction, $p<\num{0.005}$); not reliably on
larger benchmarks. The effect-map finding of
\cref{sec:results:llm} is itself the contribution: the LLM
helps where it can add qualitative architectural insight.

\section{Practical Implications for Indian Civil Engineering}
\label{sec:concl:practical}

The thesis is framed around Indian infrastructure, and the
IS~800:2007 compliance layer is the concrete bridge from
benchmark results to design-office utility. Three practical
implications follow.

First, the hybrid surrogate/FEM architecture of
\cref{sec:results:surrogate:feasibility} collapses the
\num{30}--\num{40}-minute-per-run cost of classical
optimisation into the sub-\num{20}-second range. For a design
office running parametric studies --- ``how does the optimum
shift if the live load increases by \num{15}\,\%?''
``does adding a cross-bracing change the tower weight?'' --- this
opens up interactive usage that was previously batch-only.

Second, the 72-bar rigorous-feasibility finding matters for
Indian transmission-tower practice specifically, where
transmission towers are routinely sized using commercial
software whose default constraint handling is
soft-penalty. A tower signed off at \num{380}~lb-per-tonne
equivalent under soft penalties may be carrying
\num{22}--\num{62}\,\% infeasibility in displacement; the
routine engineering decision to shift to hard-feasibility
enforcement would push the design mass up by
$\sim\num{45}\,\%$. This is a first-order economic fact that
the Indian civil-engineering community should be aware of,
independent of the specific framework developed here.

Third, the Streamlit UI exposes the whole stack in a form that
does not require Python literacy: an IIT~BHU M.Tech project
student, or a junior engineer at a consulting firm, can drive
an LLM-assisted IS~800-compliant optimisation without
writing code. The ten-second turnaround makes it usable during
a design-review meeting rather than as a background batch job.

\section{Limitations}
\label{sec:concl:limitations}

The thesis's limitations, already discussed in detail in
\cref{sec:results:threats}, are summarised here for
completeness.

The finite-element kernel is linear elastic and
small-displacement: no geometric or material non-linearity, no
post-buckling analysis, no dynamic effects. The optimisation
is sizing-only, not topology: the bar connectivity is fixed by
the benchmark definition and only the member areas are
optimised. Seed counts are small on 25-bar and 72-bar NSGA-II
(three each), and the PPO agent is trained and evaluated on a
single problem instance without transfer learning. The IS~800
compliance module uses a single column-buckling curve (curve
$a$) regardless of cross-section type, and the LLM is not
fine-tuned --- it is accessed through the public Anthropic
Claude API at temperature \num{0.3}.

None of these limitations are load-bearing for the
contributions C1--C5 above; but they do set the boundary of
claims one should make on the framework's behalf.

\section{Future Work}
\label{sec:conclusion:future}
\label{sec:concl:future}

Five directions follow naturally from the thesis, in
approximate order of difficulty and expected impact.

\paragraph{F1. Non-linear and dynamic analysis.} The FEM kernel
could be extended with geometric non-linearity (co-rotational
bar formulation) and dynamic analysis (Newmark-$\beta$ time
integration), following Bathe~\cite{bathe2014fem} and Cook et
al.~\cite{cook2002fem}. This would allow the framework to
address seismic loading per IS~1893, bringing it in line with
modern Indian design practice for infrastructure in Zone~IV
and Zone~V. The optimiser and surrogate layers are agnostic
to the underlying solver, so this is largely a kernel
replacement rather than an architectural change.

\paragraph{F2. Topology optimisation.} Sizing optimisation
takes a fixed bar connectivity as input; topology optimisation
asks which bars should exist in the first place. The
ground-structure approach of Achtziger~\cite{achtziger1999} and
the SIMP approach of Bendsoe and Sigmund~\cite{bendsoe2003topo}
could be wrapped behind the same \texttt{TrussProblem}
adapter, with the design vector augmented by binary existence
variables. The neural-surrogate architecture would need to
accommodate variable-dimension inputs --- one natural fit is a
graph neural network (GNN) that operates on the truss
adjacency, following Fey and Lenssen~\cite{fey2019gnn}, which is
listed as F3 below.

\paragraph{F3. GNN surrogate for variable-topology design.} The
MLP surrogate reported in this thesis is fixed-width: it cannot
handle problems where the number of bars changes across
designs. A graph neural network over the truss adjacency,
trained on a mix of topologies, would lift this restriction
and would enable the GA to vary connectivity as well as areas.
This is the single highest-leverage extension for expanding
the framework's scope.

\paragraph{F4. Transfer learning across benchmarks.} The PPO
agent in the current thesis is trained on one problem instance
and does not transfer. A shared observation encoder (a GNN, as
in F3) plus a policy trained across the full four-benchmark
family would directly test Zhao~\cite{zhao2021rl}'s
transferability claim: that the value of RL lies in
cross-instance generalisation. The architecture extension is
modest; the compute is a few GPU-days.

\paragraph{F5. Fine-tuned domain-specific LLM.} The LLM
warm-start in this thesis uses the public Claude API
unmodified. A next iteration would fine-tune an open-weights
model (e.g.\ Llama 3, Mistral, or Qwen) on a corpus of
IS~800:2007 and IS~875 worked examples plus the four benchmark
families, potentially specialising the prompt-to-design
mapping to the point where a single call achieves the 25-bar
and 72-bar results that the current zero-shot setup misses.
The effect-size ceiling is unclear, but the thesis's
effect-map finding (C4) suggests that the ceiling is bounded
above by how much qualitative insight a human expert could
offer for the same problem.

\paragraph{F6. Indian transmission-tower validation.}
Ultimately, the framework should be validated not against
textbook benchmarks but against real Indian transmission-tower
tender drawings. This requires a data-sharing arrangement with
Power Grid Corporation of India Limited (PGCIL) or a
consulting firm; it is the single step that would move the
framework from demonstration to deployment.

\bigskip
\noindent
All six directions above are open for a successor M.Tech or
Ph.D.\ cohort and are tracked in
\texttt{docs/ROADMAP.md} at the \texttt{phase-10-complete}
repository tag.
