\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Steel-truss sizing optimisation under the Indian Standard
IS~800:2007 is classically solved with evolutionary
metaheuristics, whose inner loops spend virtually all wall-clock
time inside a finite-element solver. This thesis develops an
integrated Python framework that composes classical
evolutionary optimisation (GA, PSO, NSGA-II) with three modern
AI layers --- a neural surrogate, a reinforcement-learning
agent, and a large-language-model warm-start designer --- into
one reproducible system, and validates the framework against
three canonical benchmarks: the 10-bar planar, 25-bar spatial,
and 72-bar spatial towers.

The framework matches the published optima of Sunar~\&
Belegundu (1991) on 10-bar to \num{5061.05}~lb against the
reported \num{5060.85}~lb (\num{0.004}~\% error), and
Schmit~\& Miura (1976) on 25-bar to \num{545.26}~lb against
\num{545.22}~lb (\num{0.008}~\% error). A three-hidden-layer
MLP surrogate trained on \num{10000} Latin-hypercube samples
achieves $R^2=\num{0.9993}$ on weight prediction and delivers
a $\num{132}\times$ wall-clock speedup, with
surrogate-in-loop optima matching FEM-in-loop optima to
$\num{0.09}\,\%$. On 72-bar we report a new methodological
finding: the literature-cited optima of Camp~\& Bichon (2004)
and Bekdas et al.\ (2015) are infeasible by \num{22}--%
\num{62}\,\% under rigorous constraint enforcement, and the
true rigorously-feasible optimum sits near $\num{549}$~lb
rather than the published $\num{380}$~lb --- a
$\num{45}\,\%$ gap that was previously undocumented in the
72-bar literature. An LLM warm-start (Claude Opus~4.7) reduces
generations-to-convergence on 10-bar by \num{36.8}~\% at
paired-Wilcoxon $p=\num{0.0046}$ across \num{40} seeds, with
the effect diminishing on larger benchmarks where no
comparable architectural insight is available to the model.
Every reported optimum passes the IS~800:2007 slenderness,
tension, compression, and serviceability-deflection checks.

The complete framework --- FEM solver, three optimisers,
surrogate, PPO agent, LLM designer, IS~800 compliance
module, FastAPI backend, and Streamlit UI --- ships as an
open-source repository at tag \texttt{phase-8b-complete}; the
Streamlit demo reproduces every headline number in under ten
seconds on a laptop CPU, with all LLM responses cached so that
the pipeline runs offline without an Anthropic API key. The
thesis thus contributes a validated, reproducible, and
Indian-design-code-compliant AI-assisted truss-design aid,
together with a methodological correction to the published
72-bar literature that the Indian civil-engineering community
should be aware of.

\clearpage
