\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

I wish to express my sincere gratitude to my supervisor,
\thesisguide, for his patient guidance, incisive technical
feedback, and steady encouragement across every phase of this
work. His insistence on rigorous constraint handling and on
reconciling every optimisation result against the Bureau of
Indian Standards IS~800:2007 clauses shaped the thesis's
methodological spine, and the 72-bar rigorous-feasibility
finding --- one of this thesis's principal contributions ---
grew directly out of that insistence.

I am thankful to the faculty and staff of the
\thesisdepartment, \thesisinstitute, for the classroom
instruction and laboratory facilities that underpin this
research. Particular thanks are due to the Structural
Engineering group for their generous loan of computing
resources during the overnight 200-bar batch runs.

I acknowledge the open-source scientific-software community
whose tools this thesis is built on: NumPy, SciPy, pymoo,
PyTorch, FastAPI, Streamlit, matplotlib, pandas, pypdf,
biblatex, siunitx, algpseudocode, and the Tectonic~LaTeX
engine. Anthropic's Claude~Opus~4.7 served as the warm-start
designer; all prompts and responses are cached in the
repository for reproducibility.

Finally, I thank my family and friends for their unstinting
support through the long evenings that a project of this scope
demands. Any errors of fact or judgement that remain in this
thesis are entirely my own.

\vspace{2cm}
\hfill \thesisauthor

\clearpage
